Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-21 WSSSPE1_report/WSSSPE1_summary_report.tex
small edit
minor grammar fix
small edits

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
be used to measure software sustainability, and though these questions might falsely lead
to yes or no answers, the complete set would determine a range of values for
sustainability. These question are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future, 
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users? 
\item Do users have a mechanism to ask questions and to learn about the element?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

@@

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is especially relevant in the modern era of social computing. In the context of
this paper, a utility software project could be defined as general purpose software
applicable to one or many generic tasks and/or enabling other software to run
by providing a suitable environment. Examples are collaborative development
framework such as \emph{GitHub} and \emph{Bitbucket}, workflow and scripting
frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and \emph{VisTrails},
and visualization frameworks such as \emph{visIT}. Often projects start with
specific, small set of requirements and users in mind but expand to provide
more generic functionality and appeal to a wider user community. Owing to its generic nature, utility software finds a wider users communities which in turn leads to a higher participation in development. Consequently, the process becomes user-driven and self sustaining. Software developed in such scenario often benefits from community driven development, deployment and maintenance best practices. For instance the \emph{Galaxy} project has significantly benefitted from user community participation in its development. It follows agile software development practices and implements standard practices such as test-driven development and socialized bug managing practices via \emph{trello}. Galaxy \emph{histories} and \emph{toolshed} offers easy community sharing of data and tools further promoting a collaborative environment.

@@

\subsection{Scientific Software}
Special purpose software that was developed specifically for a target use-case
scenario or keeping some fixed/frozen requirements in mind. Software projects pertaining to specific scientific domain often tend to be in a niche and the user community tends to be small to medium. Accuracy and optimization are some of the paramount requirements of most scientific software. Advantages of a smaller codebase and requirements include stability, ease of installation, configurability and integration into larger systems. While the software can stay stable and require relatively low maintenance, the responsibility is often on the shoulders of a very few developers. Community participation in the development tends to be low. Development is linear and simplistic with a limited scope to follow software best practices. Domain science oriented software developed to solve  specific problems, e.g., \emph{MINRES-QLP}, \emph{FASTMath}.

\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, e.g., MVAPICH. Explores uncharted territories. Highest risk. Long requirement-development-test cycles. Wide community support but tends to be handled by specialists rather then end users.

%
%\begin{itemize}
%  \item mvapich, Cactus, DUNE
%  \item gateways, github, VisIT, visTrails, PAAS
%  \item SimGrid, Swift, Galaxy
%  \item FastMath, R/qtl, MINRES-QLP
%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
%\end{itemize}
%

>>>>

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
be used to measure software sustainability, and though these questions might falsely lead
to yes or no answers, the complete set would determine a range of values for
sustainability. These questions are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future, 
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users? 
\item Do users have a mechanism to ask questions and to learn about the element?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

@@

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is especially relevant in the modern era of social computing. In the context of
this paper, a utility software project could be defined as general purpose software
applicable to one or many generic tasks and/or enabling other software to run
by providing a suitable environment. Examples are collaborative development
framework such as \emph{GitHub} and \emph{Bitbucket}, workflow and scripting
frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and \emph{VisTrails},
and visualization frameworks such as \emph{visIT}. Often projects start with
specific, small set of requirements and users in mind but expand to provide
more generic functionality and appeal to a wider user community. Owing to its generic nature, utility software finds a wider users communities which in turn leads to a higher participation in development. Consequently, the process becomes user-driven and self sustaining. Software developed in such scenario often benefits from community driven development, deployment and maintenance best practices. For instance the \emph{Galaxy} project has significantly benefitted from user community participation in its development. It follows agile software development practices and implements standard practices such as test-driven development and socialized bug managing practices via \emph{trello}. Galaxy \emph{histories} and \emph{toolshed} offers easy community sharing of data and tools further promoting a collaborative environment. The project very closely follows the guidelines described in~\cite{Carver_WSSSPE} and many from~\cite{Prlic_WSSSPE}.

@@

\subsection{Scientific Software}
Special purpose software that was developed specifically for a target use-case
scenario or keeping some fixed/frozen requirements in mind. Software projects pertaining to specific scientific domain often tend to be in a niche and the user community tends to be small to medium. Numerical accuracy and algorithmic optimization are some of the paramount requirements of most scientific software. Advantages of a smaller codebase and requirements include stability, ease of installation, configurability and integration into larger systems. While the software can stay stable and require relatively low maintenance, the responsibility is often on the shoulders of a very few developers. Community participation in the development tends to be low. Development is linear and simplistic with a limited scope to follow software best practices. Domain science oriented software developed to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. One exception is \emph{Kitware}.

\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest risk. Overlap with the utility software as it must serve as a facilitator to run existing software. Long requirement-development-test cycles. Wide community support but tends to be handled by specialists rather then end users. Promotes collaborations, one of the key ingredient of a sustainable process.
